---
title: "Calculating stratified mean biomass/abundance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calculating stratified mean biomass/abundance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, 
  fig.height = 4
)

library(diagram)
```

## Background

The `survdat` package is designed to work with the Northeast Fisheries Science Center's (NEFSC) bottom trawl surveys. The NEFSC has been conducting standardized bottom trawl surveys in the fall since 1963 and spring since 1968.  The surveys follow a stratified random design.  Fish species and several invertebrate species are enumerated on a tow by tow basis [@Azarovitz_1981]. The data are housed in the NEFSC's survey database (SVDBS) maintained by the Ecosystem Survey Branch.

The `get_survdat_data` function will query the NEFSC survey database (SVSDBS) and apply the appropriate calibration factors associated with gear or vessel differences throughout the time series.  However, end users are usually interested in more derived estimates of biomass or abundance that the tow by tow data retrieved via `get_survdat_data`. The stratified random design of the survey allows for the calculation of a stratified mean biomass or abundance for a species or group of species.

To facilitate the calculation of these stratified means, the `survdat` package has the function `calc_stratified_mean`.  This function is actually a wrapper of several intermediate functions.

```{r functions, echo = FALSE, fig.cap = "Work flow of the `calc_stratified_mean` function with associated intermediate functions."}
par(mar = rep(1, 4))
openplotmat(main = '')
pos <- coordinates(c(5, 5, 5, 5))

#Arrows
apos <- 0.75
#Boxes
text_size <- 0.7
edge      <- 0.07

#Colors
data.col <- 'grey'
fun.col <- 'light grey'
obj.col <- 'white'
big.col <- 'grey'

#Main Flow
straightarrow(from = pos[1, ] - c(0, 0.04), to = pos[2, ] - c(0.06, 0.04), 
              arr.pos = apos)
straightarrow(from = pos[8, ],  to = pos[9, ]  - c(0.04, 0), arr.pos = apos)
straightarrow(from = pos[9, ],  to = pos[10, ] - c(0.04, 0), arr.pos = apos)
straightarrow(from = pos[10, ], to = pos[15, ], arr.pos = apos)

#Strata_prep box in background
textrect(mid = pos[7, ] + c(0.095, 0),  radx = .18, rady = .32, lab = '', 
         shadow.size = 0, box.col = big.col)
text(pos[2, ][1], pos[2, ][2], 'strat_prep', cex = text_size)

#Polygon/area flow
bentarrow(from = pos[11, ] + c(0, 0.14), to = pos[12, ] + c(0, 0.01))
straightarrow(from = pos[12, ] + c(0, 0.01), to = pos[13, ] + c(-0.04, 0.01), 
              arr.pos = apos)

#Optional post stratify
bentarrow (from = pos[1, ]  - c(0, 0.06), to = pos[7, ] + c(0, 0.05),  
           path = 'H', lty = 3)
bentarrow (from = pos[11, ] + c(0, 0.16), to = pos[7, ] + c(0, 0.05),  
           path = 'H', lty = 3)
straightarrow(from = pos[7, ] + c(0, 0.05), to = pos[8, ] + c(-0.04, 0.05), 
              arr.pos = apos, lty = 3)

#Boxes
#Functions
textrect(mid = pos[10, ], radx = edge, rady = edge, lab = 'strat_mean',   
         cex = text_size, shadow.size = 0, box.col = fun.col)
textrect(mid = pos[7, ]  + c(0, 0.05), radx = edge, rady = edge, lab = 'post_strat',   
         cex = text_size, shadow.size = 0, box.col = fun.col)
textrect(mid = pos[12, ] + c(0, 0.01), radx = edge, rady = edge, lab = 'get_area',   
         cex = text_size, shadow.size = 0, box.col = fun.col)

#Data objects
textround(mid = pos[1, ] - c(0, 0.05), radx = edge - .08, rady = edge, 
          lab = 'survdat', cex = text_size, shadow.size = 0, box.col = data.col)
textround(mid = pos[11, ] + c(0, 0.15), radx = edge - .08, rady = edge, 
          lab = 'areaPolygon', cex = text_size, shadow.size = 0, box.col = data.col)

#R Objects
texthexa(mid = pos[9, ], radx = edge, rady = edge, lab = 'prepData',
          cex = text_size, shadow.size = 0, box.col = obj.col)
texthexa(mid = pos[8, ] + c(0, 0.05),  radx = edge, rady = edge, lab = 'poststratData',
          cex = text_size, shadow.size = 0, box.col = obj.col)
texthexa(mid = pos[15, ], radx = edge, rady = edge, lab = 'stratmeanData',
          cex = text_size, shadow.size = 0, box.col = obj.col)
texthexa(mid = pos[13, ] + c(0, 0.01), radx = edge, rady = edge, lab = 'strataArea',
          cex = text_size, shadow.size = 0, box.col = obj.col)

#Legend
straightarrow(from = pos[16, ] + c(0, 0.04) - c(0.08, 0), 
              to   = pos[16, ] + c(0, 0.04), lty = 1, arr.pos = 1)
straightarrow(from = pos[16, ] - c(0.08, 0.04), 
              to   = pos[16, ] - c(0, 0.04), lty = 3, arr.pos = 1)
textempty(mid = pos[16, ] + c(0, 0.04), lab = 'primary \npathway', cex = text_size)
textempty(mid = pos[16, ] - c(0, 0.04) + c(0, 0), lab = 'optional \npathway', cex = text_size)
textround(mid = pos[17, ], radx = 0.02, rady = 0.04, lab = 'data',
         cex = text_size, shadow.size = 0, box.col = data.col)
textrect(mid = pos[18, ], radx = 0.04, rady = 0.04, lab = 'function',
         cex = text_size, shadow.size = 0, box.col = fun.col)
texthexa(mid = pos[19, ], radx = 0.05, rady = 0.04, lab = 'R object',
         cex = text_size, shadow.size = 0, box.col = obj.col)
```

## `strat_prep`

The first function called is `strat_prep`.  As the name implies, `strat_prep` prepares the data set so that the stratified means can be calculated.  Nested within this function are the `post_strat` and `get_area` functions.  The `post_strat` function is used if not following the stratified design of the survey.  It is automatically called if the user inputs an `sf` object rather than the specified default `NEFSC strata`.  When this occurs, the R object `poststratData` generated by `post_strat` will replace the `survdat` data object and is ultimately passed through to the `prepData` object.  This will also turn on the `poststratFlag` which is used by the `strat_mean` function.

The rest of the prep work done by `strat_prep` is counting the number of tows per strata and the proportional weight of the strata.  This adds the columns `ntows` and `W.h` to the data set.  The number of tows (`ntows`) is simply the length of unique station records per stratum:

```{r ntows, eval = F}
# Count the number of stations in each year for each Region
  data.table::setkey(stations, YEAR, STRAT)
  stations[, ntows := length(STATION), by = key(stations)]
```

Proportional weights of the strata are based on their area.  The area of the strata are calculated using the `get_area` function which uses a Lambert Conformal Conic projection.  The `get_area` function creates a list of strata and their corresponding areas (A) which `strat_prep` uses to calculate the relative weight (W) of each strata h as:

\begin{equation}
    \label{PropArea}
    \begin{split}
    W_h = \frac{A_h}{\sum{A_h}}
\end{split}
\end{equation}

After counting the number of tows and calculating the relative weight of each strata, the resulting `prepData` is then passed to the `stratMean` function.

## `strat_mean`
