---
title: "How biomass estimates are calculated"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{How biomass estimates are calculated}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`survdat` is a collection of tools that:

* Query the NEFSC survey database (SVDBS) 
* Use the data to calculate estimates of biomass for species based on a geographic region

These data are available to qualified researchers upon request.

## Bottom Trawl Surveys

The Northeast Fisheries Science Center (NEFSC) has been conducting standardized bottom trawl surveys in the fall since 1963 and spring since 1968.  The surveys follow a stratified random design.  Fish species and several invertebrate species are enumerated on a tow by tow basis [@Azarovitz_1981]. The data are housed in the NEFSC's survey database (SVDBS) maintained by the Ecosystem Survey Branch.  

Direct pulls from the database are not advisable as there have been several gear modifications and vessel changes over the course of the time series [@Miller_2010]. `survdat` was developed as a database query that applies the appropriate calibration factors for a seamless time series since the 1960s.

## Methods

The `survdat` package can be broken down into two sections. The first pulls the raw data from SVDBS.
`survdat` identifies those research cruises associated with the seasonal bottom trawl surveys and pulls the station and biological data.  Station data includes tow identification (cruise, station, and stratum), tow location and date, as well as several environmental variables (depth, surface/bottom salinity, and surface/bottom temperature).  Stations are filtered for representativness using a station, haul, gear (SHG) code for tows prior to 2009 and a tow, operations, gear, and aquisition (TOGA) code from 2009 onward.  The codes that correspond to a representative tow (SHG <= 136 or TOGA <= 1324) are the same used by assessment biologists at the NEFSC.  Biological data includes the total biomass and abundance by species, as well as lengths and number at length.

The second section of the Survdat script applies the calibration factors.  There are four calibrartion factors applied (Table \@ref(tab:calibration)).  Calibration factors are pulled directly from SVDBS. Vessel conversions were made from either the NOAA Ship *Delaware II* or NOAA Ship *Henry Bigelow* to the NOAA Ship *Albatross IV* which was the primary vessel for most of the time series. The Albatross was decommisioned in 2009 and the Bigelow is now the primary vessel for the bottom trawl survey.

```{r calibration, eval = T, echo = F}
cal.factors <- data.frame(Name = c('Door Conversion', 'Net Conversion', 'Vessel Conversion I', 'Vessel Conversion II'),
                          Code = c('DCF', 'GCF', 'VCF', 'BCF'),
                          Applied = c('<1985', '1973 - 1981 (Spring)', 'Delaware II records', 'Henry Bigelow records'))
knitr::kable(cal.factors, booktabs = TRUE, caption = "Calibration factors for NEFSC trawl survey data")
```


The output from `survdat` is a data.table that contains all the station and biological data, corrected as noted above, from the NEFSC Spring Bottom Trawl Survey and NEFSC Fall Bottom Trawl Survey.

For the purposes of the aggregate biomass indicators, fall and spring survey data are treated separately. Additionally, all length data is dropped and species seperated by sex at the catch level are merged back together.

The data pulled from SVDBS is first post stratified into geographical regions (any user supplied geospatial shapefile) by labeling sampling stations by the geographical region they fell within (`post_strat.r`) Next, the total number of stations within each geographical region per year is counted using unique station records (`strat_prep.r`). Biomass is summed by species per year per region and divided by the appropriate station count to get the regional mean (`strat_mean.r`). Finally, the mean biomasses are summed over the defined regions (`swept_area.r`). 

## References
