---
title: "Pulling Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pulling Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(survdat)
```

Explain the differences between resulting data objects based on argument flags.

In Development


## Biological data

When using the flag `getBio = T` additional fields are pulled, namely Maturity, Age, Individual weight, stomach volume, stomach weight, and sex. Species that have lengths but no additional "biological" data are dropped from the data pull.

## Weight-Length data

Length-weight conversion coefficients are applied using values from Wigley et al. These reside in the table LENGTH_WEIGHT_COEFFICIENTS

## Length data

## Calibration factors

These reside in the table SURVAN_CONVERSION_FACTORS

Describe how they are applied

### Vessel effects

### Gear Effects

### Door Effects

