---
title: "Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(survdat)
```

## Connect to the database

Once you are granted permission to access the database and have Oracle's instant Client installed you can make a connection as follows:

```channel <- dbutils::connect_to_database(server="servername",uid="yourUsername")```

`channel` is an object inherited from the DBIConnection-class. This object is passed as an argument to functions in `survdat`

## Pull Data

Pulling the data takes up to 15 minutes to complete. see `get_survdat_data` for details. To pull the survey data with conversion corrections applied use the following:

```
data <- get_survdat_data(channel)
```

To pull biological traits of individual fish sampled on the survey cruise

```
data <- get_survdat_data(channel, bio=TRUE)
```

`data` is a list of 2 elements. The data pull and the set of sql statements made to pull the data


## Plot the shapefiles

Two shape files are included in the package, `strata.shp` and `EPU.shp`

To read and plot the `EPU.shp` file:

``` {r}
areaPolygon <- sf::st_read(dsn = system.file("extdata","EPU.shp",package="survdat"), quiet=T)
plot_shapefile(areaPolygon)

```


## Stratified Mean



## Swept area biomass

Swept area biomass is estimated for particular regions of interest. These regions are specified via shape files. 

All shapefiles are required to be an sf ([simple features](https://r-spatial.github.io/sf/index.html)) object. 


* To calculate swept area biomass for any one of these regions, say Georges Bank in the Spring:

``` calc_swept_area(data$survdat,areaPolygon=areaPolygon,areaDescription="EPU",filterByArea="GB",filterBySeason = "SPRING") ```

The `filterByArea` value should be one of the values found in the `areaDescription` field of the shapefile

* To return the values in TIDY format add the tidy=T argument:

``` calc_swept_area(data$survdat,areaPolygon=areaPolygon,areaDescription="EPU",filterByArea="GB",filterBySeason = "SPRING",tidy=T) ```

To visualize the overlap of the survey data and the regions, first define a coordinate reference system (crs) in which to project both the points and the shapefile. Convert the LAT and LON coordinates to the WGS84 reference ellipsoid (crs = 4326)

```
crs <- 4326
points <- sf::st_as_sf(data$survdat,coords=("LON","LAT"), crs=crs)
# plot data with regions
plot_data_area(points=data$survdat,polygons=area,crs)

```

