---
title: "Survey Design"
output: rmarkdown::html_vignette
bibliography: refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Survey Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(survdat)
```

The Northeast Fisheries Science Center (NEFSC) has been conducting standardized bottom trawl surveys
in the fall since 1963 and spring since 1968.  The surveys follow a stratified random design.  Fish
species and several invertebrate species are enumerated on a tow by tow basis [@Azarovitz_1981]. The data are housed in the NEFSC's survey database (SVDBS) maintained by the Ecosystem Survey Branch.  

Direct pulls from the database are not advisable as there have been several gear modifications and
vessel changes over the course of the time series [@Miller_2010]. T R package `survdat` was developed as a database query that applies the appropriate calibration factors for a seamless time series since the 1960s.


The R package `survdat` is used to pull and process the data. `survdat` identifies those research cruises associated with the seasonal bottom trawl surveys and pulls the station and biological data. Station data includes tow identification (cruise, station, and stratum), tow location and date, as well as several environmental variables (depth, surface/bottom salinity, and surface/bottom temperature). Stations are filtered using a station, haul, gear (SHG) code for tows prior to 2009 and a tow, operations, gear, and aquisition (TOGA) code from 2009 onward. The codes that correspond to a representative tow (SHG <= 136 or TOGA <= 1324) are the same used by assessment biologists at the NEFSC. Biological data includes the total biomass and abundance by species, as well as lengths and number at length.

`survdat` applies the calibration factors. There are four calibrartion factors applied (Table 64.1). Calibration factors are pulled directly from SVDBS. Vessel conversions were made from either the NOAA Ship **Delaware II** or NOAA Ship **Henry Bigelow** to the NOAA **Ship Albatross IV** which was the primary vessel for most of the time series. The Albatross was decommissioned in 2009 and the Bigelow is now the primary vessel for the bottom trawl survey.


```{r calibration, eval = T, echo = F}
cal.factors <- data.frame(Name = c('Door Conversion', 'Net Conversion', 'Vessel Conversion I', 'Vessel Conversion II'),
                          Code = c('DCF', 'GCF', 'VCF', 'BCF'),
                          Applied = c('<1985', '1973 - 1981 (Spring)', 'Delaware II records', 'Henry Bigelow records'))
kableExtra::kable(cal.factors, booktabs = TRUE,
      caption = "Calibration factors for NEFSC trawl survey data")
```

Describe a little about the survey process that pertains to the data.

eg.

Since 19xx all species landed on the survey vessel are measured for length. A subset of these are then individually weighed, sexed, assessed for stage of maturity, and undergo a stomach content analysis

At the tow level ... expanded number, expanded weight etc..


## References
